Opening up the court (surface) in tennis grand slams

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message = FALSE)
library(deuce)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(stringr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)
library(MASS)
library(ggrepel)
library(broom)
library(lme4)
library(cowplot)
library(ggpubr)

@@

The secondary data is partial point by point data for grand slam matches.  In this data set, each row is a point in a match with details on who won the point, serve speed, and whether a player had a forced or unforced error, winner, ace, or net point win.  There is also tournament info such as court surface, year, and time start.  Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.  However, these additional attributes are only available for about 1/8 of the data.   This partial data set consists of `r nrow(gs_partial)` observations (compared to the 5080 in the original grand slams).  The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.  The same trend is true for the loser rank.  This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.  Additionally, we found that no women's point by point data was recorded for 2015.

@@

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}

@@

```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter() +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament", values = tournament_colors)

@@

```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left plot) and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  # coord_flip() +
  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "bottom", legend.text=element_text(size=8),
        # legend.margin=margin(0, 0, 0, 0),
        legend.margin=margin(t = 0, unit='cm'),
        legend.box.margin=margin(-10,-10,-10,-10)) + 
  xlab("") +
  ylab("Estimate") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

```{r}
knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
             row.names = FALSE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
             caption = "\\label{tab:corr}Correlation matrix for random effects",
             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
```

@@

```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}", eval = FALSE}

g <- grid.arrange(gn, gf, gw, ncol = 3, 
              top = textGrob("Standardized Residuals vs. Predicted % Points Won",
                            gp=gpar(fontsize=20, fontfamily="serif")))

@@

Individual linear models predicting percent of points won show that Nadal was best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message = FALSE)

library(deuce)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(stringr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)
library(MASS)
library(ggrepel)
library(broom)
library(lme4)
library(cowplot)
library(ggpubr)

@@

The secondary data is partial point by point data for grand slam matches.  In this data set, each row is a point in a match with details on who won the point, serve speed, and whether a player had a forced or unforced error, winner, ace, or net point win.  There is also tournament info such as court surface, year, and time start.  Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.  However, these additional attributes are only available for about 1/8 of the data.   This partial data set consists of `r nrow(gs_partial)` observations (compared to the 5080 in the original grand slams).  The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.  The same trend is true for the loser rank.  This means that it is more likely for better players to have court by court data, which may affect our analysis for models when using this more detailed yet partial data.  Additionally, we found that no women's point by point data was recorded for 2015.

@@

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} Spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower than their ranking at other tournaments."}

@@

```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wimbledon matches tend to have more aces and fewer unforced errors than the other tournaments."}
ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter() +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament", values = tournament_colors)

@@

```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left) and player-level effects  for Williams, Federer and Nadal (right) under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  # coord_flip() +
  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "bottom", legend.text=element_text(size=8),
        # legend.margin=margin(0, 0, 0, 0),
        legend.margin=margin(t = 0, unit='cm'),
        legend.box.margin=margin(-10,-10,-10,-10)) + 
  xlab("") +
  ylab("Estimate") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

```{r, fig.pos = 'H'}
knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
             row.names = FALSE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
             caption = "\\label{tab:corr}Correlation matrix for random effects",
             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
```

@@

@@

Individual linear models predicting percent of points won show that Nadal is best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
